\vspace{-2cm} \LARGE Divergence-Free Smoothed Particle Hydrodynamics \\ \large Advanced Simulation and Visualization of Fluids in Computer Graphics

### 2017-01-12 report/paper.tex
Errors corrected in report

\begin{onecolabstract}
    \noindent This paper discusses the technique presented by Bender et. al.~\cite{bender} for simulating incompressible fluids with an efficient and stable Divergence-Free Smoothed Particle Hydrodynamics method.
    The method uses two solvers to obtain a divergence-free fluid, a divergence solver and a density solver to correct the density after the divergence has been corrected.
    This results in a stable an fast simulation of a divergence-free fluid even though the simulation is performed on the CPU without any parallelization.
    \vspace{1em}\\{\bf Index Terms:} Incompressible fluids, Divergence-free, SPH, divergence correction, density correction.\vspace{1em}
\end{onecolabstract}
]
\saythanks

@@

Listing \ref{lst:simulation} drafts the produced simulation after following Bender et al.~\cite{bender} proposed method in detail.
Upon starting the application will the particle neighbourhood $N_i$ be determined with a cell list solution explained in section \ref{section:neighbour}.
For each particle will the alpha $\alpha_i$ and density $\rho_i$ parameters computed, both of which are commonly reused variables and thus worth storing to reduce computation cost.
This is done before entering the simulation loop as initial values are required.
During the loop will all these parameters be updated as the simulation progresses.

Once in the simulation at line $7$ will first the time step be determined through the CFL condition explained in section \ref{section:timestep}.
The new velocities for each particle is then predicted through the included external forces which are propagated with ordinary euler integration.

@@

After the particle position $\vect{x_i}$ has been updated are the neighbourhoods and the $\alpha_i$ and $\rho_i$ factors recalculated to reflect the new positions.
Lastly, in line $19$ the divergence solver, much like the density solver, calculates the pressure forces to eliminate its error.
In this case the divergence error $\frac{D\rho_i}{Dt}$ is corrected to be equal to $0$.

@@

    In the equations $\frac{D\rho}{Dt}$ and $\frac{D\textbf{v}}{Dt}$ denotes the derivate of the density $\rho$ and the velocity $\textbf{v}$ respectively.
    The pressure, kinematic viscosity and body forces are denoted by $p,v$ and $\textbf{f}$ respectively.
    The second part of Equation~\ref{eq:nsDensity} $\nabla \cdot \textbf{v} = 0$ induce that the velocity field is free from divergence.
    From the continuity equation $\frac{D\rho}{Dt} = -\rho \nabla \cdot \textbf{v}$ and the divergence-free condition it follows that the partial derivate of the density with respect to the time is zero, which implies the equivalence of the equations in Equation~\ref{eq:nsDensity}.
    From this is follows that the density, in theory, must stay constant over time which is the implication of an incompressible fluid.
    In practice the divergence-free condition is not sufficient to guarantee incompressibility in simulations.
    The numerical time integrations will contain numerical errors which will cause density deviations due to the volume compressions.
    To avoid this a second condition $\rho - \rho_0 = 0$ called \textit{constant density condition} must be fullfilled~\cite{bender}.
    Section~\ref{divergenceSolver} and~\ref{densitySolver} give a deeper description of the solvers used in the project.

@@

Each rendered frame however have a constant time step in order to ensure no speed change in the playback of the rendered video.
This is decided from the user side as a desired fps parameter.
This means that the simulation loop can iterate multiple times within each frame adaptively depending on the maximum velocity.

@@

    The cell list is implemented by using a four dimensional vector where the first three dimensions are for the x, y and z coordinates and the fourth dimension is for storing the particles belonging to the cell.
    The amount of cells are decided by dividing the scene into cells of length $h$.
    The particles are then assigned to a cell by finding the cell's coordinates that it belongs to according to Equation ~\ref{eq:assignCell}, where $\vect{x}$ is the position of the particle and $\vect{s}_min$ is the lowest position of the scene space.
    If the particle moves out of its cell it is reassigned to the new cell it is inside.

@@

% Should be mentioned early since it is used in mostly of the algorithms
\subsection{Kernel} \label{kernel}
    A kernel function is used to simulate how particle-particle interactions decrease with the distance between the current particle and its neighbours.
    In SPH simulations this is an approximation of the Gaussian kernel function.
    Different kernels have been tested in previous works i.e. the poly6 kernel, the spiky kernel and the cubic spline kernel.
    According to Bender et. al.~\cite{bender} the cubic spline kernel presented by Monaghan~\cite{monaghan} was used.
    The kernel is described by Equation~\ref{eq:kernel}, where $q(x)=\frac{\left \| x \right \|}{h}$, $x$ is the distance between the current particle and a neighbour particle and $h$ is the support radius for the kernel.
    Particles further away than the support radius will not affect the current particle.

@@

    The pressure force of a particle is determined by

@@

    where the pressure gradient can be solved using the SPH formulation proposed by Ihmsen et al.~\cite{ihmsen}, Equation~\ref{eq:pressureGrad}.

@@

    Further on, the pressure forces $\vect{F_{j\leftarrow i}^p} $ that act from particle $i$ on all its neighbours $j$ is calculated through

@@

    where the stiffness parameter $\kappa_i$ is determined by Equation~\ref{stiffnessDensity}.

@@

    To conserve momentum it is required that all inner pressure forces sum up to zero such that $\vect{F_i^p} + \vect{F_{j\leftarrow i}^p} = 0$.
    A proposed by Bender et. al~\cite{bender} the solver computes the total force $\vect{F_{i,total}^p}$ as

@@

    Newton's second law $F = ma = m\frac{dv}{dt}$ implies that

@@

    The velocity of particle $i$ is thus changed to:

@@

    The divergence-free solver ensures the fluid to be free of density change over time by enforcing $\frac{D\rho}{Dt} = 0$, which is equivalent to the Navier-Stokes divergence free condition given in Equation \ref{eq:nsDensity}.
    \\ \\ Equation \ref{eq:dpdt} defines how the density change is computed.
    As in the density solver, the total pressure forces acting on a particle $i$ are computed.
    The stiffness parameter is calculated through~\ref{eq:kappaDivergence}.

@@

    The stiffness parameters needs to be determined iteratively since neighboring particles depend on each other.
    After the symmetric pressure force has been applied to the new velocity should the density change in Equation \ref{eq:dpdt} be equal or close to zero.
    If the error is too large the process will be iterated until desired result have been achieved.

>>>>

\begin{onecolabstract}
    \noindent This paper discusses the technique presented by Bender et. al.~\cite{bender} for simulating incompressible fluids with an efficient and stable Divergence-Free Smoothed Particle Hydrodynamics method.
    The method uses two solvers to obtain a divergence-free fluid, a divergence solver and a density solver to correct the density after the divergence has been corrected.
    This results in a stable and fast simulation of a divergence-free fluid even though the simulation is performed on the CPU without any parallelization.
    \vspace{1em}\\{\bf Index Terms:} Incompressible fluids, Divergence-free, SPH, divergence correction, density correction.\vspace{1em}
\end{onecolabstract}
]
\saythanks

@@

Listing \ref{lst:simulation} drafts the produced simulation after following Bender et al.~\cite{bender} proposed method in detail.
Upon starting the application the particle neighbourhood $N_i$ will be determined with a cell list solution explained in section \ref{section:neighbour}.
For each particle the alpha $\alpha_i$ and density $\rho_i$ parameters will be computed, both of which are commonly reused variables and thus worth storing to reduce computation cost.
This is done before entering the simulation loop as initial values are required.
During the loop all these parameters will be updated as the simulation progresses.

Once in the simulation at line $7$ the first time step will be determined through the CFL condition explained in section \ref{section:timestep}.
The new velocities for each particle is then predicted through the included external forces which are propagated with ordinary euler integration.

@@

After the particle position $\vect{x_i}$ has been updated the neighbourhoods and the $\alpha_i$ and $\rho_i$ factors are recalculated to reflect the new positions.
Lastly, in line $19$ the divergence solver, much like the density solver, calculates the pressure forces to eliminate its error.
In this case the divergence error $\frac{D\rho_i}{Dt}$ is corrected to be equal to $0$.

@@

    In the equations $\frac{D\rho}{Dt}$ and $\frac{D\textbf{v}}{Dt}$ denotes the derivate of the density $\rho$ and the velocity $\textbf{v}$ respectively.
    The pressure, kinematic viscosity and body forces are denoted by $p,v$ and $\textbf{f}$ respectively.
    The second part of Equation~\ref{eq:nsDensity}, $\nabla \cdot \textbf{v} = 0$, induce that the velocity field is free from divergence.
    From the continuity equation $\frac{D\rho}{Dt} = -\rho \nabla \cdot \textbf{v}$ and the divergence-free condition it follows that the partial derivate of the density with respect to the time is zero, which implies the equivalence of the equations in Equation~\ref{eq:nsDensity}.
    From this it follows that the density, in theory, must stay constant over time which is the implication of an incompressible fluid.
    In practice the divergence-free condition is not sufficient to guarantee incompressibility in simulations.
    The numerical time integrations will contain numerical errors which will cause density deviations due to the volume compressions.
    To avoid this a second condition $\rho - \rho_0 = 0$ called \textit{constant density condition} must be fullfilled~\cite{bender}.
    Section~\ref{divergenceSolver} and~\ref{densitySolver} give a deeper description of the solvers used in the project.

@@

Each rendered frame however have a constant time step in order to ensure no speed change in the playback of the rendered video.
This is decided from the user side as a desired fps parameter.
This means that the simulation loop can iterate multiple times within each frame adaptively depending on the maximum velocity of a particle.

@@

    The cell list is implemented by using a four dimensional vector where the first three dimensions are for the x, y and z coordinates and the fourth dimension is for storing the index of the particles that belong to the cell.
    The amount of cells are decided by dividing the scene into cells of length $h$.
    The particles are then assigned to a cell by finding the cell's coordinates that it belongs to according to Equation ~\ref{eq:assignCell}, where $\vect{x}$ is the position of the particle and $\vect{s}_{min}$ is the lowest position of the scene space.
    If the particle moves out of its cell it is reassigned to the new cell it is inside.

@@

% Should be mentioned early since it is used in mostly of the algorithms
\subsection{Kernel} \label{kernel}
    A kernel function is used to simulate how particle to particle interactions decrease with the distance between the current particle and its neighbours.
    In SPH simulations this is an approximation of the Gaussian kernel function.
    Different kernels have been tested in previous works i.e. the poly6 kernel, the spiky kernel and the cubic spline kernel.
    According to Bender et. al.~\cite{bender} the cubic spline kernel presented by Monaghan~\cite{monaghan} was used.
    The kernel is described by Equation~\ref{eq:kernel}, where $q(x)=\frac{\left \| x \right \|}{h}$, $x$ is the distance between the current particle and a neighbour particle and $h$ is the support radius for the kernel.
    Particles further away than the support radius will not affect the current particle.

@@

@@

    The pressure force of a particle is determined by Equation~\ref{eq:pressure}, where the pressure gradient can be solved using the SPH formulation proposed by Ihmsen et al.~\cite{ihmsen}, Equation~\ref{eq:pressureGrad}.

@@

@@

    Further on, the pressure forces $\vect{F_{j\leftarrow i}^p} $ that act from particle $i$ on all its neighbours $j$ is calculated through Equation~\ref{eq:pressure2}, where the stiffness parameter $\kappa_i$ is determined by Equation~\ref{stiffnessDensity}.

@@

    To conserve momentum it is required that all inner pressure forces sum up to zero such that $\vect{F_i^p} + \vect{F_{j\leftarrow i}^p} = 0$.
    As proposed by Bender et. al~\cite{bender} the solver computes the total force $\vect{F_{i,total}^p}$ through Equation~\ref{eq:pressureTotal}.

@@

    Newton's second law $F = ma = m\frac{dv}{dt}$ implies that $\frac{dv}{dt}$ is according to Equation~\ref{dvdt}. 
    The velocity of particle $i$ is thus changed according to Equation~\ref{eq:velocity}.

@@

@@

    The divergence-free solver ensures the fluid to be free of density change over time by enforcing $\frac{D\rho}{Dt} = 0$, which is equivalent to the Navier-Stokes divergence free condition given in Equation \ref{eq:nsDensity}.
    \\ \\ Equation \ref{eq:dpdt} defines how the density change is computed.
    As in the density solver, the total pressure forces acting on a particle $i$ are computed.
    The stiffness parameter is calculated through Equation~\ref{eq:kappaDivergence}.

@@

    The stiffness parameters needs to be determined iteratively since neighboring particles depend on each other.
    After the symmetric pressure force has been applied to the new velocity the density change in Equation \ref{eq:dpdt} should be equal or close to zero.
    If the error is too large the process will be iterated until desired result have been achieved.
